---
title: "Reproducible Research Project 1"
author: "Dan Piltzer"
output: md_document
---
```{r loading packages and settings, echo = FALSE, results='hide', include=FALSE}
library(dplyr)
library(ggplot2)
options(scipen=999)
```

###1. Loading and Pre-processing the Data

```{r Loading and Pre-processing the Data, results='hide'}
data <- read.csv("activity.csv")
complete_data <- na.omit(data)
```
  
  
###2. What is mean total number of steps taken per day?
```{r mean total number of steps taken per day}
total_steps <- sum(complete_data$steps)
data_by_day <- group_by(complete_data, date) %>% summarize(total_steps = sum(steps))
hist(data_by_day$total_steps, main = "Steps per Day", xlab = "Steps per Day")
mean_steps <- round(mean(data_by_day$total_steps))
median_steps <- median(data_by_day$total_steps)
```
The mean steps taken per day is `r mean_steps` and the median steps taken per day is `r median_steps`. The total steps taken during the time period is `r total_steps`.
  
  
###3. What is the average daily activity pattern?
```{r}
interval_data <- group_by(complete_data, interval) %>% summarize(average_steps = mean(steps))  
ggplot(interval_data, aes(interval, average_steps))+geom_line()+labs(title = "Average steps by 5 minute interval")
max_interval <- interval_data$interval[which.max(interval_data$average_steps)]
```
The interval with the largest average number of steps is `r max_interval`.
  
  
###4. Imputing missing values
```{r}
num_NA <- sum(is.na(data$steps))
completer_data <- merge(data, interval_data, "interval")
completer_data$steps[is.na(completer_data$steps)] <- completer_data$average_steps[is.na(completer_data$steps)]
completer_day_data <- select(completer_data, -average_steps) %>% group_by(date) %>% summarize(total_steps = sum(steps))

hist(completer_day_data$total_steps, main = "Steps per Day", xlab = "Steps per Day")

mean_steps2 <- round(mean(completer_day_data$total_steps))
median_steps2 <- round(median(completer_day_data$total_steps))
```
This time the mean steps taken per day is `r mean_steps2` which is the same as last time.  The median steps per day is `r median_steps2` which is now different than the last time.  
  
  
###5. Are there differences in activity patterns between weekdays and weekends?
```{r}
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday","Saturday", "Sunday")
day_group <- c("Weekday", "Weekday", "Weekday", "Weekday", "Weekday", "Weekend", "Weekend")
weekends <- cbind(weekday, day_group)
completer_data <- mutate(completer_data, weekday = weekdays(as.Date(date))) %>% merge(weekends, "weekday") %>% select(-weekday,-average_steps) %>% group_by(interval,day_group) %>% summarize(average_steps = mean(steps))
ggplot(completer_data, aes(interval, average_steps)) + geom_line() + facet_grid(day_group ~.) + labs(title = "Average steps by 5 minute interval")

```
